---
title: "Hands-on - Exploring data with R"
author: "Dr. Marc A.T. Teunis; Dr. Jan-Willem Lankhaar"
date: "12 September 2019"
subtitle: 'Borrelsessie IM/ICT'
always_allow_html: yes
output: github_document
resource_files:
- data/heights.csv
- data/heights_outliers.csv
- images/hex-ggplot2.png
- code/ggqq.R
- code/08_gganimate_demo.R
---

_From R. A. Fisher, 1938: To consult the statistician after an experiment is finished is often merely to ask him to conduct a post mortem examination. He can perhaps say what the experiment died of_

## Attribution 
This work can be reproduced under licence CC BY-NC, assuming you give attribution to the original creators. See authors of this document and the 'resources' document in the Github repo for more details.

```{r setup, include=FALSE}
## knitr global settings
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE, 
                      fig.width = 5, fig.height = 3.5)

ggplot2::theme_set(ggplot2::theme_bw())

```

## Loading the packages used in this demo
```{r}
library(tidyverse)
library(gganimate)
library(dslabs)
library(lubridate)
library(ggplot2)
library(gapminder)
library(gganimate)
library(gifski)
library(tidyverse)

```

## Measles example
Every data cycle starts with a question / problem:

 - Problem
 - Plan
 - Data
 - Analysis
 - Conclusion

```{r}
 DiagrammeR::grViz("
digraph rmarkdown {
  Problem -> Plan
  Plan -> Data
  Data -> Analysis
  Analysis -> Conclusion 
  Conclusion -> Problem
}
")

```
 
## Was the introduction of the vaccination against the measles effective in the USA?
```{r}
library(dslabs)
data(us_contagious_diseases)
the_disease <- "Measles"
dat <- us_contagious_diseases %>%
  dplyr::filter(!state%in%c("Hawaii","Alaska") &  
      disease == the_disease) %>%
  mutate(rate = count / population * 10000 * 52 / weeks_reporting) 

jet.colors <- colorRampPalette(c("#F0FFFF", "cyan", "#007FFF", "yellow", "#FFBF00", "orange", "red", "#7F0000"), bias = 2.25)

dat %>% mutate(state = reorder(state, desc(state))) %>%
  ggplot(aes(year, state, fill = rate)) +
  geom_tile(color = "white", size = 0.35) +
  scale_x_continuous(expand = c(0,0)) +
  scale_fill_gradientn(colors = jet.colors(16), na.value = 'white') +
  geom_vline(xintercept = 1963, col = "black") +
  theme_minimal() + 
  theme(panel.grid = element_blank()) +
  coord_cartesian(clip = 'off') +
  ggtitle(the_disease) +
  ylab("") +
  xlab("") +  
  theme(legend.position = "bottom", text = element_text(size = 8)) + 
  annotate(geom = "text", x = 1963, y = 50.5, label = "Vaccine introduced", size = 3, hjust = 0)


```

## Build upon existing code

Try changing the line in the code chunk above from

`the_disease <- "Measles"` (_Mazelen_ in Dutch)  to

`the_disease <- "Smallpox"` (_Pokken_ in Dutch) and rerun the code.

Can you find out which part of the code above plots the vertical line and prints the words "Vaccine introduced". Try to adapt the figure for the Smallpox data: The introduction of the smallpox vaccine was already as early as in 1940. **TIP** You will need to change the year 1963 twice to 1940 in order for the new figure to be correct for the Smallpox subset of the data. Try adapting the code and rerun the code chunk.

Question: When did vaccination against smallpox stop? Try googling for the answer! 

## Exploring the Gapminder dataset in R

The gapminder dataset contains a number of measurements on health and income outcomes for 184 countries from 1960 to 2016. It also includes two character vectors, OECD and OPEC, with the names of OECD and OPEC countries from 2016. 

## Inspecting the dataset with R
```{r}
data("gapminder", package = "dslabs")
head(gapminder)
gapminder <- gapminder %>% as_tibble()
names(gapminder)
```

## A very simple example to start with
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point()
```

# Fixing overplotting
We call this overlay of points 'overplotting'.

This can be fixed in several ways:

 - Reducing the transparency of data points  
 - Mapping colour to a variable (continuous or categorical)
 - Reduce the data in the plot
 - Mapping a shape to a variable
 - Add noise (`"jitter"`) to points
 - Choosing different `geom_`
 - Facetting; create panels for 'categorical' or so-called 'factor' variables in R
 - Summarize the data
 - Displaying a model / relationship that represents the data (and not sho the actual data itself) 
 - Or any combination of the above strategies

__Basically you map an `aesthetic` (`aes()`) or a `characteristic` to a variable__
 
Let's go over a few of these overplotting methods  

### 1. Overplotting: Reducing transparency (`alpha`) of points or lines in the data
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(alpha = 0.1)
```

### 2. Mapping colour to a variable (continuous or categorical)
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = continent))
```

#### or combined with `alpha`
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.1) +
  guides(colour = guide_legend(override.aes = list(alpha = 1)))

```

#### Choose a different `geom_`
```{r}
# install.packages("ggpointdensity")
library(ggpointdensity)

gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  stat_density2d(aes(fill = continent), geom = 'polygon', alpha = 0.5) +
  guides(fill = guide_legend(override.aes = list(alpha = 1))) 
  
```


#### **Do it yourself:**

 - Try adjusting some of the arguments in the previous `ggplot2` call. 
For example, adjust the `alpha = ...` or change the variable in `x = ...`, `y = ...` or `colour = ...`
 - `names(gapminder)` gives you the variable names that you can change
 - Show and discuss the resulting plot with your neighbour
 - What do you think this part does: 
 
 `guides(... = guide_legend(override.aes = list(alpha = 1)))`
 
 - Try to find out by disabling with `#`

### 3. Reduce the data in the plot
```{r}
reduce_data_plot <- gapminder %>% 
  filter(continent == "Africa" | continent == "Europe") %>%
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.2) +
  ## override the alpha setting for the points in the legend:
  guides(colour = guide_legend(override.aes = list(alpha = 1))) 

reduce_data_plot
```

### **Discuss with you neighbour:**

 - What does the the `aes()` part of the `geom_point()` do?
 
### 4. Mapping a shape to a variable
```{r}
## or e.g. show only two years and map a shape to continent
shape_plot <- gapminder %>% 
  dplyr::filter(continent == "Africa" | continent == "Europe",
         year == "1960" | year == "2010") %>%
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = as_factor(as.character(year)), 
                 shape = continent), 
             alpha = 0.7)

shape_plot
```

#### **Do it youself**

 - Try removing the `as_factor(as.character(year))` call and replace this by only `year` above and rerun the plot, what happened?

### 5. Facetting

Create panels for 'categorical' or so-called 'factor' variables in R
```{r}
facets_plot <- gapminder %>% 
  dplyr::filter(continent == "Africa" | continent == "Europe",
         year == "1960" | year == "2010") %>%
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.5) +
  facet_wrap(~ year)

facets_plot
```

### 6. Summarize the data
```{r}
library(ggrepel)

years <- c("1960", "1970", "1980", "1990", "2000", "2010")

summarize_plot <- gapminder %>% 
  dplyr::filter(year %in% years) %>%
  group_by(continent, year) %>%
  summarise(mean_life_expectancy = mean(life_expectancy),
            mean_fertility = mean(fertility)) %>%
  ggplot(aes(x = mean_fertility,
             y = mean_life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.7) 

summarize_plot
```

#### Adding labels to points with `{ggrepel}`
```{r}
library(ggrepel)

years <- c("1960", "1970", "1980", "1990", "2000", "2010")

labels_plot <- gapminder %>% 
  dplyr::filter(year %in% years) %>%
  group_by(continent, year) %>%
  summarise(mean_life_expectancy = mean(life_expectancy),
            mean_fertility = mean(fertility)) %>%
  ggplot(aes(x = mean_fertility,
             y = mean_life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.7) +
  geom_label_repel(aes(label=year), size = 2.5, box.padding = .5)
  

labels_plot
```

### 7. Displaying a model / relationship that represents the data (and not show the actual data itself) 
```{r}
## Model
lm <- gapminder %>% lm(formula = life_expectancy ~ fertility)

correlation <- cor.test(x = gapminder$fertility, y = gapminder$life_expectancy, method = "pearson")

# save predictions of the model in the new data frame 
# together with variable you want to plot against
predicted_df <- data.frame(gapminder_pred = predict(lm, gapminder), 
                           fertility = gapminder$fertility)

```

#### Add model to plot
```{r}
model_plot <- gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(alpha = 0.03) +
  geom_line(data = predicted_df, aes(x = fertility, 
                                     y = gapminder_pred),
            colour = "darkred", size = 1)
model_plot
```

#### Plotting statistics to the graph with the `{ggpubr}` package
```{r, echo=FALSE}
library(ggpubr)
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(alpha = 0.02) +
  
  geom_line(data = predicted_df, 
            aes(x = fertility, 
                y = gapminder_pred),
            colour = "darkred", size = 1) +
  stat_cor(method = "pearson", label.x = 2, label.y = 30) +
  theme_bw()

```

#### Using a smoother `geom_smooth` to display potential relationships
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(alpha = 0.02) +
  geom_smooth(method = "lm") +
  stat_cor(method = "pearson", label.x = 2, label.y = 30) +
  theme_bw()
```


# Time series
So far we have been mapping colours and shapes to categorical variables. You can also map to continuous variables though.
```{r}
continuous <- gapminder %>%
  dplyr::filter(country == "Netherlands" | 
                country == "China" |
                country == "India") %>%
  dplyr::filter(year %in% years) %>%
  ggplot(aes(x = year,
         y = life_expectancy)) +
  geom_point(aes(size = population, colour = country)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  geom_line(aes(group = country)) +
  theme_bw()

continuous
```

## Summarize per continent and sum population
```{r}
population_plot <- gapminder %>% 
  dplyr::group_by(continent, year) %>%
  dplyr::filter(year %in% years) %>%
  summarise(sum_population = sum(population)) %>% 
  ggplot(aes(x = year, 
             y = sum_population)) +
    geom_point(aes(colour = continent)) +
    geom_line(aes(group = continent,
                  colour = continent))
population_plot
```

# Filtering for one country
```{r}
dslabs::gapminder %>%
  dplyr::filter(country == "Rwanda") %>%
  ggplot(aes(x = year, y = life_expectancy)) +
  geom_point() +
  geom_line() +
  ggtitle("Life expectancy for Rwanda over the years") 
  
```


# Ranking data
```{r}
ranking_plot <- gapminder %>%
  dplyr::filter(continent == "Europe",
                year == 2010) %>%
  ggplot(aes(x = reorder(as_factor(country), population),
             y = log10(population))) +
  geom_point() +
  ylab("log10(Population)") +
  xlab("Country") + 
  coord_flip() +
  geom_point(data = filter(gapminder %>%
  dplyr::filter(continent == "Europe",
                year == 2010), population >= 1e7), colour = "red")

ranking_plot
```

# Let's look at a time series
We filter for "Americas" and "Oceania" and look at `life_expectancy` over the years.
```{r}
## without summarizing for countries
gapminder$continent %>% as_factor() %>% levels()
gapminder %>% 
  dplyr::filter(continent == "Americas" | continent == "Oceania") %>%
  ggplot(aes(x = year,
             y = life_expectancy)) +
  geom_line(aes(group = continent,
                colour = continent))
```  

**Obviously something went wrong here. Please, discuss with your neighbour what you think happened or needs to be done to fix this (without looking ahead ;-) )**
























## Grouping
We can see what happened if we plot individual datapoints
```{r}
gapminder %>% 
  dplyr::filter(continent == "Americas" | continent == "Oceania") %>%
  ggplot(aes(x = year,
             y = life_expectancy)) +
  geom_point(aes(colour = country)) +
  theme(legend.position="none") +
  facet_wrap( ~ continent) +
  theme(legend.position="none") 
  
```

## Summarizing time series data
```{r}
gapminder$continent %>% as_factor() %>% levels()
gapminder %>% 
  dplyr::filter(continent == "Americas" | continent == "Oceania") %>%
  group_by(continent, year) %>%
  summarise(mean_life_expectancy = mean(life_expectancy)) %>%
  ggplot(aes(x = year,
             y = mean_life_expectancy)) +
  geom_line(aes(group = continent,
                colour = continent)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Statistical proof?
```{r}

df <- gapminder %>% 
  dplyr::filter(continent == "Americas" | continent == "Oceania") %>%
  group_by(continent, year)

model <- aov(data = df, life_expectancy ~ continent * year)
anova(model)

```

Some remarks on the above Two-way ANOVA:

 - Repeated measures / multilevel models might be more appropriate here (paired / nested designs)
 - We did not perform any check on assumptions
 - We performed our analysis on only part of the data
 
## One more option: categorical values and "jitter"
Sometimes you have overlapping plots and adding transparency with `alpha()` or mapping colour to underlying categorical values is not working because there are simple to many points overlapping

Let's look at an example
```{r}
gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  group_by(continent) %>%
  dplyr::filter(year %in% years) %>%
  ggplot(aes(x = year,
             y = infant_mortality)) +
  geom_point(aes(colour = country)) +
  theme(legend.position="none")
  
```

In such cases it can be helpfull to add some noise to the points (`position = "jitter"`) to reduce overlapping. This can be a powerfull approach, especially when combined with setting `alpha()`
```{r}
gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
    group_by(continent) %>%
  ggplot(aes(x = year,
             y = infant_mortality)) +
  geom_point(aes(colour = continent), position = "jitter") 
  
```
 
## Bar chart 
It would be nice to know what the mean child mortality is for both continents
```{r}
gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
  group_by(continent, year) %>%
  summarise(mean_infant_mortality = mean(infant_mortality, na.rm = TRUE)) %>% 
  ggplot(aes(x = year,
             y = mean_infant_mortality)) +
  geom_col(aes(fill = continent), position = "dodge") 
  
```

## Adding summary data to an existing plot
Now that we have the mean infant mortality for each year for the two continents, let's add that data to the previous dot plot where we used jitter
```{r}
mean_inf_mort <- gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
  group_by(continent, year) %>%
  summarise(mean_infant_mortality = mean(infant_mortality, na.rm = TRUE))

gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
    group_by(continent) %>%
  ggplot(aes(x = year,
             y = infant_mortality)) +
  geom_point(aes(colour = continent), position = "jitter") +

## summary data added to previous 
  geom_line(data = mean_inf_mort, aes(x = year, 
                                      y = mean_infant_mortality, 
                                      colour = continent),  size = 2)


```

## Filter data from a graph
In the figure above we can observe a number of countries in 'Americas' continent that have a child mortality that are above the average (over the years) of 'Africa'. Which countries are this?

```{r}
library(ggiraph)

gapminder$country <- 
  str_replace_all(string = gapminder$country, 
                pattern = "'", 
                replacement = "_")


interactive_inf_mort <- gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
    group_by(region, country) %>%
  ggplot(aes(x = year,
             y = infant_mortality)) +
  
  geom_point_interactive(aes(tooltip = country, colour = region), position = "jitter") +
  
#  geom_point(aes(colour = continent), position = "jitter") +

## summary data added to previous 
 geom_line(data = mean_inf_mort, aes(x = year, 
                                      y = mean_infant_mortality, 
                                      colour = continent, group = continent),  size = 2
            )

interactive_inf_mort

gapminder$country %>% as_factor() %>% levels()
ggiraph(ggobj = interactive_inf_mort)

```






 
# A more complicated example (for showing the capabilities of ggplot2)
```{r}
west <- c("Western Europe","Northern Europe","Southern Europe",
          "Northern America","Australia and New Zealand")

gapminder <- gapminder %>%
  mutate(group = case_when(
    region %in% west ~ "The West",
    region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
    region %in% c("Caribbean", "Central America", "South America") ~ "Latin America",
    continent == "Africa" & region != "Northern Africa" ~ "Sub-Saharan Africa",
    TRUE ~ "Others"))
gapminder <- gapminder %>%
  mutate(group = factor(group, levels = rev(c("Others", "Latin America", "East Asia","Sub-Saharan Africa", "The West"))))

filter(gapminder, year%in%c(1962, 2013) & !is.na(group) &
         !is.na(fertility) & !is.na(life_expectancy)) %>%
  mutate(population_in_millions = population/10^6) %>%
  ggplot( aes(fertility, y=life_expectancy, col = group, size = population_in_millions)) +
  geom_point(alpha = 0.8) +
  guides(size=FALSE) +
  theme(plot.title = element_blank(), legend.title = element_blank()) +
  coord_cartesian(ylim = c(30, 85)) +
  xlab("Fertility rate (births per woman)") +
  ylab("Life Expectancy") +
  geom_text(aes(x=7, y=82, label=year), cex=12, color="grey") +
  facet_grid(. ~ year) +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank(),
        strip.text.y = element_blank(),
   legend.position = "top")
```


# Data Distributions & Outliers

## Detecting outliers
For this part we use a different and more simple dataset
This dataset contains 1192 observations on self-reported:

 - `height` (inch)
 - `earn` ($)
 - `sex` (gender)
 - `ed` (currently unannotated)
 - `age` (years)
 - `race` 
 
```{r}
heights_data <- read_csv(file = here::here("data",
                                          "heights_outliers.csv"))

heights_data
```

## Data characteristics
We will focus on the variable `height` here
```{r}
summary_heights_data <- heights_data %>%
  group_by(sex, age) %>%
  summarise(mean_height = mean(height, na.rm = TRUE),
            min_height = min(height),
            max_height = max(height)) %>%
  arrange(desc(mean_height))

summary_heights_data[c(1:4),]
```

From the above summary we can conclude that there are two outliers (presumably entry errors).

**Calculate the height in meters for each outlier in the `Console`** 1 inch = 0,0254 meters

**Please discuss the solution with your neighbour** 

## Checking the frequency distribution
```{r}

heights_data %>%
  ggplot(aes(x = height)) +
  geom_histogram(aes(stat = "identity"), bins = 200)
```

This distribution looks odd. 
When you see a large x-axis with no data plotted on it, it usually means there is an outlier. If you look carefully, you will spot two outliers around 600

## Boxplots to detect outliers
```{r}
heights_data %>%
  ggplot(aes(y = height)) +
  geom_boxplot()
```

So apparantly there is one data point that is way off from the rest of the distribution. Let's remove this point, using `filter()` from the `{dplyr}` package like we did before on the `gapminder` dataset.
```{r}
heights_data %>%
  dplyr::filter(height < 100) %>%
  ggplot(aes(y = height)) +
  geom_boxplot()

## by sex

heights_data %>%
  dplyr::filter(height < 100) %>%
  ggplot(aes(y = height, x = sex)) +
  geom_boxplot()

```

## New frequency distribution
Now let's plot a new distribution plot, this time we plot density, leaving the outlier out
```{r}
heights_data %>%
  dplyr::filter(height < 100) %>%
  ggplot(aes(height)) +
  geom_freqpoly(aes(y = ..density..))

## by sex
heights_data %>%
  dplyr::filter(height < 100) %>%
  ggplot(aes(height)) +
  geom_freqpoly(aes(y = ..density.., colour = sex))

```

## A formal check to detect an outlier - handle with care
The best way to spot and take care of an outlier (**you never delete an outlier unsubstantiated!!**) is to create multiple visualizations as decribed above. However, there is a formal, simple statistical apporach to assess whether an observation is an outlier.
The apporach below can only be used to detect a single outlier in a numeric vector. For more complex (multivariate) situations see e.g.: http://r-statistics.co/Outlier-Treatment-With-R.html 

### The Dixon's Q Test for detecting outliers
The `{outliers}` package contains a number of tools. Here we will demonstrate the `outlier()` and `dixon.test()` functions. 
For more information on the DQT, see: https://www.statisticshowto.datasciencecentral.com/dixons-q-test/

### `outlier()` function
This function finds the value with largest difference between it and sample mean, which can be an outlier.

We will walk trough a simple exercise that first shows the result of the `outlier()` function and than we calculate it by hand

First we create dummy data for a triplicate series of measurements. We assume 5 measurement values, replicated three times, containing one outlier. We also assume is was a concentration depended series in which the index indicates the used concentration.
```{r}
set.seed(123)
## create data
random_numbers <- rnorm(n = 15, mean = 2.5, sd = 0.9) 
## inject outlier
random_numbers[4] <- 9.54
## define triplicates
triplicates <- rep(1:3, times = 5)
index <- rep(1:5, times = 3)

## put in dataframe
measured_data <- tibble(measured = random_numbers, replicate = triplicates,
                        index = index)
measured_data

## sort data
measured_data %>% 
  arrange(desc(measured))
```

## Plotting the data reveals the outlier.
```{r}
## plot
measured_data %>% 
#  enframe() %>%
  ggplot(aes(x = index, y = measured)) +
  geom_point() +
  geom_point(data = dplyr::filter(.data = measured_data, measured > 5), 
             colour = "red", size = 6, shape = 4)
```

## Identify with `outliers::outlier()`
```{r}
outliers::outlier(measured_data$measured)
```

### Identify with `outliers::dixon.test()`
```{r}
## note that these functions from the {outliers} package take a vector (use $)
outliers::dixon.test(measured_data$measured)
```
The test is significant against the NULL hypothesis that the highest value is not an outlier. We can conclude that this NULL hypothesis does not hold on the basis of the significant p-value < 0.05

### Manual calculations:

*`outlier()`*

```{r}
mean_vctr <- mean(measured_data$measured)

distance <- measured_data$measured - mean_vctr

ind <- distance == max(distance)

measured_data[ind,]

```

*`dixon.test()`*

The Dixon's Q Test can be calculated manually following the following procedure:

How to Run Dixon’s Q Test (R10).

Note: make sure your data set is normally distributed before running the test; for example, run a Shapiro-Wilk test. Running it on different distributions will lead to erroneous results. An extreme value may throw off any test for normality, so try running that test without the suspect data item. If your data set still doesn’t meet the assumption of normality after running a test for it, then you should not run Dixon’s Q Test. 

Note that for very small sample sizes the assumtion for Normal distribution usually does not hold. We will see more formal way of establishing Normal distribution below. 

**Caution: the test should not be used more than once for the same set of data.**


To perform the calculations:

Step 1: Sort your data into ascending order (smallest to largest).

Step 2 :Find the Q statistic using the following formula:

$Qexp = \frac{x2 - x1}{xn - x1}$

Where 

 * $x1$ is the smallest (suspect) value
 * $x2$ is the second smallest value 
 * $xn$ is the largest value


Dixon’s Q Test: Definition, Step by Step Examples + Q Critical Values Tables

Find Outliers > Dixon’s Q Test
What is Dixon’s Q Test?

Dixon’s Q test, or just the “Q Test” is a way to find outliers in very small, normally distributed, data sets. Small data sets are usually defined as somewhere between 3 and 7 items. It’s commonly used in chemistry, where data sets sometimes include one suspect observation that’s much lower or much higher than the other values. Keeping an outlier in data affects calculations like the mean and standard deviation, so true outliers should be removed.

Dixon came up with many different equations to find true outliers. The most commonly used one is called the R10 or simply the “Q” version, which is used to test if one single value is an outlier in a sample size of between 3 and 7. Dean and Dixon did suggest various other formulas in a later paper, but these are not commonly used. For a full list of alternate formulas for different sample sizes (up to about 30), go to: Dixon’s Test, Alternate Formulas and Tables.
How to Run Dixon’s Q Test (R10).

Note: make sure your data set is normally distributed before running the test; for example, run a Shapiro-Wilk test. Running it on different distributions will lead to erroneous results. An extreme value may throw off any test for normality, so try running that test without the suspect data item. If your data set still doesn’t meet the assumption of normality after running a test for it, then you should not run Dixon’s Q Test,

Caution: the test should not be used more than once for the same set of data.

Step 1: Sort your data into ascending order (smallest to largest).
167, 177, 180, 181, 185, 188, 189.

Step 2 :Find the Q statistic using the following formula:
dixon's q test statistic

Where:

    x1 is the smallest (suspect) value,
    x2 is the second smallest value,
    and xn is the largest value.

Step 3: Find the Q critical value in the Q table (see below). For a sample size of 3 and an alpha level of 5%, the critical value is 0.970.

Step 4: Compare the Q statistic from Step 2 with the Q critical value in Step 3. If the Q statistic is greater than the Q critical value, the point is an outlier.

Let's try it on our `measured_data`

```{r}
measured_data_ranked <- measured_data %>%
  arrange(measured) 
measured_data_ranked$measured

x1 <- measured_data_ranked$measured[1]
x2 <- measured_data_ranked$measured[2]
xn <- max(measured_data_ranked$measured)

Qexp = (x2-x1) / (xn-x1)

##
0.5/8

```






## Checking normality with a qqplot
a qqplot provides a visual aid to assess whether a distribution is approaching normality
```{r}
## 
source(file = here::here("code", "ggqq.R"))
height_data_outlier_removed <- heights_data %>%
  dplyr::filter(height < 100)
  gg_qq(height_data_outlier_removed$height) 
  
## formal statistical proof
shapiro.test(height_data_outlier_removed$height)
```

all data -> reject hypothesis that the sample has a normal distribution

## Test individual distributions
```{r}
males <- height_data_outlier_removed %>%
  dplyr::filter(sex == "male")

females <- height_data_outlier_removed %>%
  dplyr::filter(sex == "female")

shapiro.test(males$height)
shapiro.test(females$height)
## add shapiro for each sex

## we can do the same for age
shapiro.test(males$age)
shapiro.test(females$age)
```
